---
title: "EDA for Networks"
output: html_document
date: "2023-10-06"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, message = FALSE}
## Set directories
data_path <- 'Data/'
save_path <- 'Data/'

# Load packages
library(tidyverse)
library(here)
library(igraph)
library(ggraph)
library(RColorBrewer)
library(pheatmap)
library(nimble)
library(nimbleHMC)
library(bayesplot)

# Do we want to save any output?
save_files <- FALSE

```

# Network data structures

To introduce some of the most useful techniques for working with network
data, we begin with an example based on trade flows involving OECD
countries in 2021.

Note: the terms "vertex", "node", "country" can and will used
interchangeably in this example, as can "edge",
"link","connection","trade flow."

## Load and explore the data

Network data is typically provided as a data frame in which the first
two columns contain identifiers between which an edge is observed, and
additional columns define edge attributes. Alternately, data may be
provided as an edge list, a two column matrix containing only the vertex
identifiers.

Our OECD data is stored as a data frame with supplementary information
in the accompanying file `tradecountries.txt`. We load both in the code
below.

```{r load_data}
# Load the files
file_path1 <- here::here(paste0(data_path, "2021Trade/trade2021.txt"))
trade <- read.table(file_path1, sep = '\t',header = TRUE, skip = 0 ,quote='', comment='')

file_path2 <- here::here(paste0(data_path, "2021Trade/tradecountries.txt"))
trade.countries <- read.csv(file_path2, sep = '\t',header = TRUE, skip = 0 ,quote='', comment='')
```

Browsing the data, we see that it is provided as a directed, weighted
edge list with direction corresponding to the flow from exporter to
importer and weight corresponding to the value of the trade flow.
Additionally, we notice that we have more recipient ("to") countries
than reporting countries ("from). Also note that since trade flows of
any size are reported, this is a highly dense network.

```{r}
head(trade)

# Note we have more to countries than from and only from countries are in our country list
unique(trade$from)
unique(trade$to)
trade.countries

# Consider the proportion of edges realized
nC.from <- length(unique(trade$from))
nC.to <- length(unique(trade$to))
nC.from*(nC.to - 1)
nrow(trade)
```

To begin visualizing the network, we use the default plotting function of the
`igraph` package - don't worry if it doesn't look great, we'll improve the 
aesthetics later!

```{r}
# Convert the data frame into an igraph object and view
g <- graph_from_data_frame(trade, directed = TRUE)
plot(g) 
```

## Subset and transform the data

Next, we simplify our data from a very dense, weighted, directed network
to a sparse weighted, undirected network. The most important question is
how to meaningfully define an edge. In this case, let's define an edge
to indicate a strong mutual trade relationship: an edge exists between
countries A and B if A sends at least 1% of its exports to B and B also
sends at least 1% of its exports to B.

To compute this, we'll first need to compute the total value of exports
from each country in the list and restrict our data to countries for
which we have both import and export data (OECD countries).

```{r, cleaning1}

# Restrict dataset to countries with both imports and exports (i.e., OECD countries)
trade.plus <- trade[trade$to %in% trade$from, ] %>% # restrict countries
  group_by(from) %>% 
  mutate(amount_total_from = sum(amount)) %>% # create variable to record total exports
  mutate(prop_from = amount/amount_total_from ) %>% 
  ungroup() %>% 
  group_by(to) %>% 
  mutate(amount_total_to = sum(amount)) %>% # create variable to record total imports
  mutate(prop_to = amount/amount_total_to)

# Add a new useful variable: our network will be undirected
# Replace to-from columns with undirected vertex list (alphabetized)
trade.plus$country1 <- apply(trade.plus, 1, function(x) sort(unlist(c(x[1], x[2])))[1])  
trade.plus$country2 <- apply(trade.plus, 1, function(x) sort(unlist(c(x[1], x[2])))[2])  

# Define a new edge type by thresholding the proportional trade values
trade.subset <- trade.plus %>% 
                filter(prop_from > 0.01 & prop_to > 0.01) %>%
                group_by(country1, country2) %>% 
                mutate(trade_volume = sum(amount)) %>% 
                select(country1, country2, trade_volume) %>% 
                unique()

nrow(trade.subset) # much more sparse!
```

While the simplified network above will be our primary data set, we can preserve
additional vertex features in a separate data frame.
```{r}
# Preserve the node-specific vars above in the vertex df
trade.countries.subset <- trade.countries %>% 
        filter(code %in% union(unique(trade.subset$country1 ), unique(trade.subset$country2))) %>%
        inner_join(., trade.plus, by = c("code" = "from")) %>%
        select(code, name, amount_total_from) %>%
        unique() %>% 
        mutate(region = ifelse(code == "AUS" | code == "NZL", "Oceania", 
                        ifelse(code == "CHL" , "South America", 
                        ifelse(code == "ISR" | code == "TUR", "Middle East",
                        ifelse(code == "USA" | code == "MEX" | code == "CAN", "North America",
                        ifelse(code == "JPN", "Asia", "Europe"))))))


```

# Visualization

## Graphs

Now that we have a more sparse network, we can visualize it
as a graph with nodes representing countries, and edges representing a
strong mutual trade relationshiop as defined above.

We begin by experimenting with different graph layouts, including the
force-directed Fruchterman-Reingold algorithm.

```{r, vis, fig.height = 10}
# Convert our subset data frame to an igraph object
g <- graph_from_data_frame(trade.subset, directed = FALSE, vertices = trade.countries.subset)

# Experiment with layouts
par(mfrow = c(2,2))
plot(g, main = "Default") # Force directed
plot(g, layout = layout.circle, main = "Circle")
plot(g, layout = layout.sphere, main = "Sphere")
plot(g, layout = layout.fruchterman.reingold, main = "Fruchterman Reingold")
par(mfrow = c(1,1))
```

Next, we can add edge weight, representing the total volume of trade
between two countries, and node size, representing the total amount
exported from each country. Both attributes should be scaled *roughly*
to a (0,1) interval. We experiment with a few transformations below. 


```{r}
# Set edge and scale weights
E(g)$weight <- trade.subset$trade_volume
E(g)$weight <- E(g)$weight / max(E(g)$weight) # scale by max
E(g)$weight <- log(trade.subset$trade_volume) # log volume
E(g)$weight <- as.numeric(scale(trade.subset$trade_volume, center = FALSE)) # scale by sd, as.numeric removes the scaling attribute

# Set node size
V(g)$size <- trade.countries.subset$amount_total_from
V(g)$size <- V(g)$size/max(V(g)$size)
V(g)$size <- log(trade.countries.subset$amount_total_from)

# Color and labels
V(g)$name <- trade.countries.subset$code
V(g)$color <- as.factor(trade.countries.subset$region)

k1 <- 5
k2 <- 5
plot(g, edge.width = E(g)$weight/5, vertex.size = V(g)$size/5)
plot(g, edge.width = E(g)$weight/k1, vertex.size = V(g)$size/k2) 
```


While node size and edge width appear to be reasonable scaled, the nodes still appear too
small, and there is too much text overlap. While we can produce
effective graphs using the `igraph` package, `ggraph` offers greater
control of graph aesthetics. Basic usage is demonstrated below, and a
useful tutorial is
[here](https://www.mr.schochastics.net/material/netVizR/).

```{r}
# Pass the igraph object created above to ggraph
ggraph(g, layout = 'stress') + 
  geom_edge_link(aes(width = weight), alpha = 0.25) + # Set edge weight with attributes of g
  scale_edge_width(range = c(0.5, 2.5)) +  # Constrain edge width
  geom_node_point(aes(size = size, color = color), alpha = 0.5) + # Set node color with attributes of g
  scale_size(range = c(2,10)) + # Constrain node sizes
  geom_node_text(aes(label = name), repel = TRUE, point.padding = unit(0.5, "lines")) + 
  theme_void() + 
  theme(legend.position = "none")

```

## Adjacency matrices and heatmaps

While there are many ways to express network data, the most useful
mathematical representation is through an *adjacency matrix*: a binary
adjacency matrix $A$ has entries $A_{ij} = 1$ if node $i$ and node $j$
have a link and 0 otherwise. While we will mostly be working with binary
adjaceny matrices this week, note that the components could also
indicate counts indicating the number of links between nodes, or weights
indicating the intensity of the link.

We begin by extracting the adjacency matrix from our graph object.

```{r}
# Create adjacency matrix and define number of countries 
A.dgC <- as_adjacency_matrix(g) # Gives us a dgC matrix object
A <- as.matrix(A.dgC) # simple numeric array
nC <- nrow(trade.countries.subset)
```

The adjacency matrix easily lends itself to the heatmap visualization in
which the values of our adjacency matrix are depicted with different
colors. Typically, the rows of an adjacency matrix (nodes) are reordered
by a clustering algorithm to produce a more effective visualization.

We can also provide an annotation dataframe which can be used to label the rows and columns. 

```{r}
# Create annotation dataframe
df.an <- trade.countries.subset %>% # annotation df must have same rownames as A
         remove_rownames() %>%
         column_to_rownames(var = "code") %>%
         select(region)

# Basic heatmap: nice, but I always forget if red means high or low
pheatmap(A)

# Annotated heatmap: grayscale
pheatmap(A, cluster_rows = T, cluster_cols = T,
         color=colorRampPalette(brewer.pal(9,"Greys")[c(1,8)])(30), 
         main = "Observed Adjacency Matrix", 
         fontsize = 8, show_rownames = T, show_colnames = F, 
         legend = F, annotation_col = df.an)
```

------------------------------------------------------------------------

### Exercise

Create a heatmap using edge weights for the color value. Does the clustering changerelative to the binary heatmap created above?

------------------------------------------------------------------------


## Plotting the spectral decomposition

One additional visualization involves representing nodes by their
corresponding components in the leading two eigenvectors. Because our
graph is undirected, its adjacency matrix is symmetric (and real) and so
it is orthogonally diagonalizable and its eigenvalues are real.

```{r}
# Take spectral decomposition of the unweighted adjacency matrix
eig.A <- eigen(A)

# Plot posterior clustering
evec.A <- eig.A$vectors

data.frame(V1 = evec.A[,1], V2 = evec.A[,2], Region = as.factor(trade.countries.subset$region), 
                 size = rowSums(A), code = trade.countries.subset$code) %>% 
  ggplot(aes(x = V1, y = V2, color = Region, size = size, label = code), alpha = 0.25) + 
  #geom_jitter() + 
  geom_text(check_overlap = TRUE, show.legend = FALSE) + 
  theme_minimal() + 
  guides(size = "none") + 
  labs(title = "Data") + 
  scale_color_viridis_d()

```

In the plot above, text size corresponds represents the degree of each
node, what do you notice about the relationship between degree and the
first eigenvector?

# Network and vertex statistics

## Vertex metrics

As suggested above, the spectral decomposition leads to a useful measure
of the overall influence of a node or vertex in the network, known as
the *eigenvector centrality*. Recall the other key vertex metrics discussed in lecture:

-   *Eigenvector centrality*: $x_i = 1/\lambda \sum_{j \in N(i)} x_i = 1/\lambda \sum_{j \in V} a_{i,j} x_j$, 
    where $x_i$ is the centrality score of node $i$; this implies that a node's
    centrality socre is proportional to the sum of the centrality scores of its neighbors; measure of a node's influence in a network
-   *Degree*: $d_i = \sum_{j \neq i} y_{ij}$ is the number of edges
    between node $i$ and other nodes
-   *Betweenness centrality*: the sum of proportions of shortest paths
    that pass through a given vertex. Let $\sigma_{jk}$ denote the total
    number of shortest paths between nodes $j,k$ and $\sigma_{jk} (i)$
    denote the number of shortest paths that pass through $i$. Then the
    betweenness centrality of node $i$ is
    $\sum_{i \neq j \neq k}\frac{\sigma_{jk}(v)}{\sigma_{jk}}$. Note
    that betweenness as defined above scales with the number of node
    pairs and so it is useful to normalize by dividing by the number of
    (undirected) pairs excluding $i$, $(N-1)(N-2)/2$
-   *Closeness centrality*: inverse sum of shortest path lengths
    $s_{ij}$ to every other node from node $i$ (only defined for
    connected graphs): $\frac{1}{\sum_{j \neq i} s_{ij}}$. We can
    normalize by instead taking the inverse of the average shortest path
    length.


To compute these in R:
```{r}

# Degree: most basic property, number of edges adjacent to a vertex
degree <- rowSums(A)

# Betweenness
betweenness(g, v = V(g), directed = FALSE, weights = NA)
betweenness(g, v = V(g), directed = FALSE, weights = NA, normalized = TRUE)

# Closeness 
closeness(g, vids = V(g), weights = NA)
closeness(g, vids = V(g), weights = NA, normalized = TRUE)

# Eigen centrality
eigen_centrality(g, weights = NA)$vector
eigen(A)$vectors[,1]

```

------------------------------------------------------------------------

### Exercise

Comment on degree distribution. Does this network have hubs? 

Verify that eigencentrality is proportional to the sum of neighbors' eigencentrality for 
one or two nodes. 

------------------------------------------------------------------------

## Network metrics

While vertex summaries measure the influence of any given vertex in the
network, network metrics provide a variety of insights into the overall
structure of a network.

### Clustering coefficients

The first network-level summary we'll consider is the clustering
coefficient, which measures the tendency of nodes to cluster. Let
$\mathcal{N}_i = \{j: y_{ij} =1\}$ denote the set of all nodes connected
to node $i$. In the code below we calculate the following clustering
statistic:

-   *Local clustering coefficient*: the proportion of possible links in
    $\mathcal{N}_i$ that are realized:
    $C_i = \frac{2} {|\mathcal{N}_i|(|\mathcal{N}_i|-1)}\sum_{j:y_{ij}=1,k:y_{ik}=1}y_{jk}$
-   *Global clustering coefficient*: the proportion of all triplets that
    are closed
    $C = \frac{\sum_{i,j,k} y_{ij}y_{ik}y_{jk}}{\sum_i |\mathcal{N}_i|(|\mathcal{N}_i|-1)}$
-   *Network average clustering coefficient*:
    $\overline{C}=(1/n)\sum_{i=1}^n C_i$

```{r}

# Local clustering coefficient
C.i <- transitivity(g, type = "localundirected")
cbind(trade.countries.subset$name, round(C.i, 2))

# Global clustering coefficient
C <- transitivity(g, type = "globalundirected")
C

# Network average clustering coefficient
C.bar <- transitivity(g, type = "localaverageundirected")
C.bar
mean(C.i, na.rm = TRUE)
```

### Shortest paths and small world networks

Random graphs generated from the Erdos-Renyi model exhibit small
clustering coefficients and small average shortest path length.

Small world networks: the distance $L$ between any two randomly chosen
nodes grows proportionally to $log(N)$, where $N$ is the number of nodes
in the network. Small world networks are characterized by high values of
clustering coefficient and low average path length between nodes, with

$L \propto log(N)$

```{r}
sps <- shortest.paths(g, v = V(g), to = V(g), weights = NA)
mean(sps[upper.tri(sps)])

```

------------------------------------------------------------------------

### Exercise

Does it appear that our trade network displays the small world property? 
Does our network resemble and Erdos-Renyi random graph?

```{r}
g.sw <- sample_smallworld(dim = 1, size = 100, nei = 3, loops = FALSE, multiple = FALSE, p = 0.05)
sps <- shortest.paths(g.sw, v = V(g.sw), to = V(g.sw), weights = NA)
hist(sps)


```

------------------------------------------------------------------------

### Other network stats.

-   *Diameter*: length of the longest shortest path
-   *Density*: proportion of all possible edges that are realized
-   *Modularity*: measures the tendency of a graph to cluster with
    respect to an observed node membership variable

```{r}

# Connectance aka density: number of observed interactions divided by the number of possible interactions
sum(A[lower.tri(A)])/(nC*(nC -1)/2)
edge_density(g)

# Diameter
diameter(g, directed = FALSE, weights = NA)
max(sps)

# Modularity
modularity(g, membership = as.numeric(as.factor(trade.countries.subset$region)), weights = NULL)

```

------------------------------------------------------------------------

### Exercise

Characterize the trade network with respect to the vertex and network
metrics described above. Is the network structure robust to different
edge definitions and subsets?

------------------------------------------------------------------------

### Save the clean data

Save the clean, subset network for use in subsequent exercises.

```{r save}

if(save_files){
  write.csv(A, here::here(paste0(data_path, "2021Trade/A_subset.csv")), row.names = FALSE)
  write.csv(trade.countries.subset, here::here(paste0(data_path, "2021Trade/Features_subset.csv")), row.names = FALSE)
}
```

------------------------------------------------------------------------

### Exercise

***

Throughout the course of the week you will be analyzing a data set of
your choice. Selct a data set from among:

-   The example data sets provided in Data folder of the course Github
    repo (excluding of course the small trade dataset that we have used
    in this exercise)

-   The various network data repositories shared in the Course resources
    document

-   Any network data sets you have encountered in your work and research

Perform EDA on your selected network, starting with the techniques
above. Because the majority of models discussed in this course will
apply to undirected graphs, select a data set where an undirected edge
can meaningfully be defined. Additionally, think about potential edge
covariates.
